---
output:
  html_document: default
  pdf_document: default
---
Red Wine Data Exploration by Christian Guzman
========================================================

```{r echo=FALSE, message=FALSE, warning=FALSE, packages}

library(ggplot2)
library(gridExtra)
library(reshape2)
library(RColorBrewer)
```

```{r echo=FALSE, Load_the_Data}
# Clear workspace
rm(list = ls())
# Load the Data
wine <- read.csv("wineQualityReds.csv")
```

> This report explores a data set containing quality rating and other attributes 
for approximately 1,600 red wines.

# Univariate Plots Section

```{r echo=FALSE, Univariate_Plots}
list(summary(wine), names(wine))


print(str(wine))

```

As we can see by the structure of the data set, ther are 13 features, with all
the features being of numeric type, except for the dependent variable, "quality" 
which is a integer type.

```{r echo=FALSE, Outputvariable}
qplot(quality, data = wine, binwidth = 1) +
  scale_x_continuous(breaks = seq(3, 8, 1))
```

```{r echo=FALSE}
summary(wine$quality)
table(wine$quality)
```

The first plot is a histogram showing the distribution of the quality.

Since there are few wines rated below 5, and above 7 the data will barely be 
noticeable. As such I will combine the quality ratings.

```{r echo=FALSE, creating.rating.feature}
wine$rating <- cut(wine$quality,
                           breaks = c(2, 5, 6, 8), 
                           labels = c("Poor", "Good", "Premium"))
summary(wine$rating)
```

Wines rated from 3 to 5, 6, and 7 to 8 will be labeled as "Poor", "Good", and
"Premium" respectively.

```{r echo=FALSE}
typeof(wine$rating)
str(wine$rating)
wine$rating <- as.factor(wine$rating)

```

The new feature "rating" is a factor of 3 levels. We can use this new feature
to analyze the data in new ways.

```{r echo=FALSE, bar.graph}
qplot(wine$rating)
```

The bar graph shows each level with a significant amount of wines. Premium
happens to have significantly fewer data that Poor and Good wines.

```{r echo=FALSE, grid.arrange, message=FALSE}
with(data = wine, {
  grid.arrange(qplot(fixed.acidity) +
                 scale_x_continuous(breaks = seq(4, 16, 2)),
               qplot(volatile.acidity),
               qplot(citric.acid),
               qplot(residual.sugar),
               qplot(chlorides),
               qplot(density),
               qplot(pH),
               qplot(alcohol),
               qplot(free.sulfur.dioxide),
               qplot(total.sulfur.dioxide),
               ncol = 2)
})
```

I would describe the majority of the plots to be skewed.
Specifically, all variables except for "pH" and "density" have obvious right 
skew. With "residual.sugar", "chlorides" free and total sulf dioxide as having 
strong right skew, and long right tails.

```{r echo=FALSE, fixed.acdity}
qplot(wine$fixed.acidity, fill = I('#5b0000'), color = I('black'),
      binwidth = .2) +
  scale_x_continuous(breaks = seq(4, 16, 2)) +
  xlab("Fixed Acidity") +
  ylab("Frequency")
```

Fixed acidity (Tartaric acid): We can see that most wines have fixed acidity
value between 6 and 9 g/L.

```{r echo=FALSE}
qplot(wine$fixed.acidity, fill = I('#5b0000'), color = I('black'),
      binwidth = .020) +
  scale_x_log10(breaks = c(4.6, seq(4, 16, 2))) +
  xlab("Fixed Acidity (g/L)") +
  ylab("Frequency")
by(data = wine$fixed.acidity, INDICES = wine$rating, FUN = summary)
```

The above histogram shows the log transformation of the "fixed.acidity" 
variable. The log transformation can help view the data with much less skew and
with usually less extreme outliers. The log transform will be applied to most of 
the data.

Dubbed the most important process in winemaking, I hope to discover its 
relationship to other features and wine quality.

```{r echo=FALSE, message=FALSE}
grid.arrange(qplot(data = wine, x = volatile.acidity),
      qplot(data = wine, x = volatile.acidity, color = I('black'), 
      fill = I('#5b0000')) +
  scale_x_log10(breaks = c(seq(.1200, .32, .05), seq(.32, 1.5800, .17))))

summary(wine$volatile.acidity)
by(data = wine$volatile.acidity, INDICES = wine$rating, FUN = summary)
```


Volatile acidity (acetic acid): The transformed distribution is unimodal, and
peaks near between .49 and 0.66 g/L. I would imagine this feature being an 
important variable for quality since too high a level can give the wine an 
unpleasant vinegar taste.

```{r echo=FALSE, citric.acid, message=FALSE}
grid.arrange(qplot(data = wine, x = citric.acid),
      qplot(data = wine, x = citric.acid +.1, color = I('black'), 
      fill = I('#5b0000')) +
  scale_x_log10(breaks = seq(0, 1, .1)))

summary(wine$citric.acid)
by(data = wine$citric.acid, INDICES = wine$rating, FUN = summary)

```

Citric acid: The distribution has a right skew, and the transformed  
distribution appears bimodal with a peak being near 0.6 and the other peak near
0.1. Many of the values are clustered near 0.

Another observation to note is that higher quality wines have a higher mean and
median for citric acid. This could imply that citric acid is an import factor in 
wine quality especially since it gives wine its freshness.

```{r log_transformation, echo=FALSE, message=FALSE}
# Standard summary
summary(wine$residual.sugar)
# log10 transformation summary
summary(log10(wine$residual.sugar + .1))



grid.arrange(ggplot(aes(x = residual.sugar), data = wine) + 
  geom_histogram() + scale_x_continuous(breaks = seq(0, 15.5,
                                                                    1)),
  ggplot(aes(x = residual.sugar), data = wine) + 
  geom_histogram(color = 'black', 
                 binwidth = .05, fill=I('#5b0000')) +
  scale_x_log10(breaks = c(0, 0.9, 1.5, 2, 2.5, 3,  seq(4, 15.5, 2)))
  )

```

For the dataset we can see that most of the wines have between 1.5 and 3.0 grams 
per liter of residual sugar.

```{r echo=FALSE}
summary(wine$residual.sugar)
by(data = wine$residual.sugar, INDICES = wine$rating, FUN = summary)
```

As can be seen by the summary of residual sugar content by wine rating, there is 
no clear trend. Could it be that residual sugar is not an important feature?
The lower rated wines do have a higher max value but similar medians, and 
premium wines have a higher mean of residual sugar content.

```{r echo=FALSE}
summary(log10(wine$chlorides))
u <- median(wine$chlorides)

grid.arrange(
  qplot(data = wine, x = chlorides),
  qplot(data = wine, x = chlorides, color = I('black'), 
        fill = I('#5b0000')) +
    scale_x_log10(breaks = c(0, min(wine$chlorides),
                                    (u - min(wine$chlorides)) / 2,
                                    u,
                                    seq(u, (u + max(wine$chlorides)) / 2,
                                        .07),
                                    max(wine$chlorides)))
  )
```

The chlorides variable is the other strongly right skewed variable. With a 
log transformation we can get a better sense of the structure of the data.

The distribution is unimodal with the median being 0.0790 grams/liter of salt 
content in the wines. Most values are between .0335 and .1490.

```{r echo=FALSE, density}
grid.arrange(
  qplot(x = density, data = wine),
  qplot(x = pH, data = wine))
summary(wine$density)
```

The distributions for density and pH resemble a normal distribution.

```{r echo=FALSE, alcohol, message=FALSE}
grid.arrange(
  qplot(x = alcohol, data = wine),
  qplot(x = alcohol, data = wine, color = I('black'), fill = I('#5b0000')) +
    scale_x_log10(breaks = seq(0, 14, .5)))
summary(wine$alcohol)
table(as.factor(round(wine$alcohol)))
```

As can be seen a log transformation hardly transforms the distribution. We can 
see More clearly however that alcohol content peaks at 9.5.

```{r faceting, echo=FALSE, message=FALSE}
qplot(x = alcohol, data = wine) +
  scale_x_continuous()+
  facet_wrap(~rating, ncol = 1)
```

Plotting the distribution of alcohol faceting on wine quality rating variable,
a trend can be seen that there are far less higher rated wines with lower 
alcohol content and the poor rated wines peak at 9.5. The distributions shift to
the right.

```{r echo=FALSE}
by(data = wine$alcohol, INDICES = wine$rating, FUN = summary)
```

```{r echo=FALSE}
qplot(x = alcohol, data = subset(wine, rating == "Premium"), 
      geom = "histogram", binwidth = .25, color = I('black'))
summary(wine$alcohol[wine$rating == "Premium" ])
```

From the above the premium wines group have a median of 11.60 and good and poor 
wine groups have medians 10.50 and 9.70 respectively indicating a trend that
higher rated wines are more likely to have higher alcohol content.

```{r sulfur.dioxide, eval=FALSE, message=FALSE}
qplot(x = free.sulfur.dioxide / total.sulfur.dioxide, data = wine)
with(wine,{
     summary(free.sulfur.dioxide / total.sulfur.dioxide)}
     )
```

I wanted to get a look at free sulfur dioxide divided by total sulfur dioxide
because I felt this proportion value would give more information. I know from
online research that wine makers strive for a certain percentage of free sulfur
dioxide to combat bacterial growth.

From the above summary, the min percentage was approximately 2% with the median
being 37.5% and the max was 86%. It would be interesting to discover what is the
prefered percentage of free total sulfur dioxide for the higher rated wines.


# Univariate Analysis

### What is the structure of your dataset?
The dataset contains 12 features, with 1,599 observations. One extra feature 
'X' is simply an index and was removed from dataset. The "sulphates" feature is
also excluded from this analysis. "rating" feature was created from making 
"quality" into a 3 level factor.

```{r echo=FALSE, remove_feature_X}
wine$X <- NULL
wine$sulphates <- NULL
names(wine)
```

```{r Locate missing data if any, echo=TRUE}
length(wine[!complete.cases(wine)])
```

There was no missing data.

### What is/are the main feature(s) of interest in your dataset?

The main feature of interest is the created feature "rating". The attempt is to
explore what features factor into the decision of the experts to rate a wine 
high or low.

### What other features in the dataset do you think will help support your \
investigation into your feature(s) of interest?

The main features of interest for me are those that have obvious affect on the 
taste of the wine such as the acidity features (fixed acidity, volatile acidity, 
citric acid), residual sugar, density and alcohol, PH,  free and total sulfur dioxide. 

### Did you create any new variables from existing variables in the dataset?

I created the variable "rating".

### Of the features you investigated, were there any unusual distributions? \
Did you perform any operations on the data to tidy, adjust, or change the form \
of the data? If so, why did you do this?

I applied log transformations for the heavily right skewed data such as: 
fixed acidity, volatile acidity, citric acid, and residual sugar. I did this
to lower the influence of outliers in order to better view the data. I also made
quality into a nominal variable of 3 levels. This is important because I have to
take into account that my dependent variable is categorical.



# Bivariate Plots Section

```{r echo=FALSE}
mosthighlycorrelated <- function(mydataframe,numtoreport)
{
  # find the correlations
  cormatrix <- cor(mydataframe)
  # set the correlations on the diagonal or lower triangle to zero,
  # so they will not be reported as the highest ones:
  diag(cormatrix) <- 0
  cormatrix[lower.tri(cormatrix)] <- 0
  # flatten the matrix into a dataframe for easy sorting
  fm <- as.data.frame(as.table(cormatrix))
  # assign human-friendly names
  names(fm) <- c("First.Variable", "Second.Variable","Correlation")
  # sort and print the top n correlations
  head(fm[order(abs(fm$Correlation),decreasing=T),],n=numtoreport)
}

attr(mosthighlycorrelated, "comment") <- "Prints out linear correlation coefficients for each pair of variables in data set, in order of the correlation coefficient.
Ref: https://little-book-of-r-for-multivariate-analysis.readthedocs.io/en/latest/src/multivariateanalysis.html#plotting-multivariate-data"

mosthighlycorrelated(wine[1:9], 6)
```

These are the top 10 most highly correlated pairs of features. From this table
we can say that fixed acidity is strongly correlated with PH, citric acid, and
density. 

Free sulfur dioxide is strongly correlated with total sulfur dioxide
with a positive pearson score of 0.67. This implies that free sulfur dioxide
accounts for nearly 70% of the variance in total sulfur dioxide and there is a
positive trend that as free sulfur dioxide increases so to does the total 
amount.



```{r correlation.matrix2, echo=FALSE, fig.width=14.5, fig.height=8}
library(GGally)
library(ggiraphExtra)
# Ref: http://jamesmarquezportfolio.com/correlation_matrices_in_r.html
ggcorr(wine, nbreaks=8, palette='RdGy', label=TRUE, label_size=5, label_color='white')
```

```{r echo=FALSE}
cor.test(wine$alcohol, wine$quality)
cor.test(wine$volatile.acidity, wine$quality)
```

The wine quality independent variable has strongest positive and negative
pearson score with alcohol and volatile acidity respectively.

```{r echo=FALSE}
grid.arrange(qplot(x = alcohol, y = quality, data = wine),
             qplot(x = volatile.acidity, y = quality, data = wine))
```

Despite the fact that quality has integer values we can barely make out the 
positive and negative slope by the fact that as alcohol increases there are more
higher rated wines, and as volatile acidity increases there are more lower 
quality wines.

```{r echo=FALSE}
grid.arrange(
ggplot(aes(y = alcohol, x = as.factor(quality)), data = wine) +
  geom_boxplot() + coord_flip(),
ggplot(aes(y = alcohol, x = rating), data = wine) +
  geom_boxplot() + coord_flip()
)
by(data = wine$alcohol, INDICES = wine$rating, FUN = summary)
```

Using a box plot the trends are easier to observe when making quality into a
categorical variable. The box plot supports the observation that as alcohol 
increases so to does quality score.

```{r box.plot, echo=FALSE}
grid.arrange(
ggplot(aes(y = volatile.acidity, x = as.factor(quality)), data = wine) +
  geom_boxplot() + coord_flip(),
ggplot(aes(y = volatile.acidity, x = rating), data = wine) +
  geom_boxplot() + coord_flip()
)
by(data = wine$volatile.acidity, INDICES = wine$rating, FUN = summary)
```

A similar observation can be made with volatile acidity but instead the medians 
of the higher rated groups are lower, indicating a negative relationship. 

```{r eval=TRUE, echo=FALSE, correlation_matrix, message=FALSE, fig.width=14.5, fig.height=8}
library(GGally)

theme_set(theme_minimal(20))
set.seed(1836)
wine_subset <- wine[, c(1:3,8:10,12)]
names(wine_subset)
ggpairs(wine_subset[sample.int(nrow(wine), 1000),], mapping = aes(color = rating),
        columns = c(4:7))

```

The scatter and density plots I will go over in more detail in the multivariate
portion of the analysis.

```{r acids_v_density, echo=FALSE}
qplot(data = wine, x = fixed.acidity, y = pH)
qplot(data = wine, x = fixed.acidity, y = citric.acid)
qplot(data = wine, x = fixed.acidity, y = density)
```

The strongest correlation found was between fixed acidity and PH. This is an
obvious one because pH is a scale used to acidity. In this case the more tartaric 
acid in the wine the lower pH it has. I expect to see similar results for the 
other acids (volatile acidty, citric acid).

```{r echo=FALSE}
ggplot(data = wine, aes(x = fixed.acidity, y = pH)) +
  geom_jitter(alpha = 1/4, position = position_jitter(h=0)) +
  coord_cartesian(xlim = c(5, 12)) +
  geom_smooth(method='lm')
```

Using jitter and alpha we can remedy the problem of overplotting.

```{r echo=FALSE}
qplot(data = wine, x = volatile.acidity, y = pH)
cor(x = wine$volatile.acidity, y = wine$pH)
```

The pearson score and visualization in the above suggests a weaker linear 
relationship. More volatile acidity could decrease the acidity.

```{r echo=FALSE}
ggplot(data = wine, aes(x = volatile.acidity, y = pH)) +
  geom_jitter(alpha = 1/2, position = position_jitter(h=0)) +
  coord_cartesian(xlim = c(0.2, 0.8)) +
  geom_smooth(method='lm')
```

Using jitter and alpha transparency and zooming settings in we can make out 
where most of the points are falling, and make out a trend. 

```{r echo=FALSE, citric.v.pH}
qplot(data = wine, x = citric.acid, y = pH)
cor(x = wine$citric.acid, y = wine$pH)
```

While citric acid and pH has a pearson score of -0.54. Citric acid contributes 
to over 50% of the variance in pH and like fixed acidity, the more of this acid 
the lower the pH suggesting higher acidity.

```{r echo=FALSE}
ggplot(data = wine, aes(x = citric.acid, y = pH)) +
  geom_jitter(alpha = 1/3.5, position = position_jitter(h=0)) +
  coord_cartesian(xlim = c(0, 0.5)) +
  geom_smooth(method='lm')
```

Applying the jitter and alpha transperancy I can see three clusters
where the data points are heavily concentrated in. There is also a vertical band 
on the left indicating wines with 0 citric acid have varying pH values.

```{r echo=FALSE}
qplot(data = wine, x = volatile.acidity, y = citric.acid)
cor(x = wine$volatile.acidity, y = wine$citric.acid)
```

Another strong correlation was between volatile acidity and citric acid. It is 
interesting to note that if fixed acidity tends to increase citric acid and if 
volatile acidity decreases citric acid this could be the same as saying that 
fixed acidity improves the "freshness" of the wine, while too much volatile 
acidity tends to decrease it.

```{r echo=FALSE}
ggplot(data = wine, aes(x = volatile.acidity, y = citric.acid)) +
  geom_jitter(alpha = 1/3, position = position_jitter(h=0)) +
  coord_cartesian(xlim = c(0.25, 0.9)) +
  geom_smooth(method='lm')
```

The visualization does appear to show a vague downward slope as volatile acidity 
increases.

```{r echo=FALSE}
ggplot(aes(y = citric.acid, x = rating), data = wine) +
  geom_boxplot() + coord_flip()
```

More citric acid seems to be an important aspect to premium wines with a clear
trend for higher ratings having more of it.

```{r echo=FALSE}
cor(x = wine$fixed.acidity, y = wine$quality)
ggplot(aes(y = fixed.acidity, 
           x = as.factor(quality)), data = wine) +
  geom_boxplot() + coord_flip()

ggplot(aes(y = fixed.acidity, 
           x = rating), data = wine) +
  geom_boxplot() + coord_flip()
cor(x = wine$fixed.acidity, y = wine$quality)
by(data = wine$fixed.acidity, INDICES = wine$rating, FUN = summary)
```

With the box plots there is no clear visible trend.

```{r echo=FALSE}
ggplot(aes(y = residual.sugar, x = rating), data = wine) +
  geom_boxplot() + coord_flip() +
  coord_cartesian(ylim = c(1,3.5))
```

Residual sugar also does not show strong correlation with quality rating, as I 
thought it would initially.

```{r echo=FALSE}
ggplot(aes(y = free.sulfur.dioxide / total.sulfur.dioxide, 
           x = rating), 
       data = wine) +
  geom_boxplot() + coord_flip()
```


# Bivariate Analysis

### Talk about some of the relationships you observed in this part of the \
investigation. How did the feature(s) of interest vary with other features in \
the dataset?

The main feature of interest, quality was most strongly correlated with alcohol 
and volatile acidity.

There was no clear visible trend for fixed acidity and rating nor was there for 
residual sugar and rating.

Better wines tend to have higher proportion of free sulfur dioxide.

### Did you observe any interesting relationships between the other features \
(not the main feature(s) of interest)?

As fixed acidity content increases citric acid and density also tend
to increase. But as fixed acidity increases pH value tends to decrease, 
indicating an increase in acidity.

Citric acid is strongly correlated with fixed acidity and volatile acidity. The 
pearson scores are 0.67 and -0.55 respectively.


### What was the strongest relationship you found?

Fixed acidity was negatively and strongly correlated with pH, positively 
correlated with citric acid and density.

free sulfur dioxide was also strongly and positively correlated with total 
sulfur dioxide but this is to be expected since free SO2 is a portion of total 
SO2.

From this exploration and analysis I might not want to use fixed acidity with 
pH, citric acid, or density to reduce colinearity between the explanatory 
variables to better understand the relationships between the independent 
variable and the dependent variables.

# Multivariate Plots Section

```{r density.plots, echo=FALSE}
ggplot(aes(x = free.sulfur.dioxide / total.sulfur.dioxide), data = wine) +
  geom_density(aes(color = rating), size = 1) +
  scale_color_brewer(type = 'seq', palette = "YlGnBu") +
  theme(panel.background = element_rect(fill = 'dark gray'))

ggplot(aes(x = fixed.acidity), data = wine) +
  geom_density(aes(color = rating), size = 1) +
  scale_color_brewer(type = 'seq', palette = "YlOrBr") +
  theme(panel.background = element_rect(fill = 'dark gray'))

ggplot(aes(x = volatile.acidity), data = wine) +
  geom_density(aes(color = rating), size = 1) +
  scale_color_brewer(type = 'seq', palette = "YlGn") +
  theme(panel.background = element_rect(fill = 'dark gray'))

ggplot(aes(x = citric.acid), data = wine) +
  geom_density(aes(color = rating), size = 1) +
  scale_color_brewer(type = 'seq', palette = "RdPu") +
  theme(panel.background = element_rect(fill = 'dark gray'))

ggplot(aes(x = pH), data = wine) +
  geom_density(aes(color = rating), size = 1) +
  scale_color_brewer(type = 'seq', palette = "PuBuGn") +
  theme(panel.background = element_rect(fill = 'dark gray'))

ggplot(aes(x = alcohol), data = wine) +
  geom_density(aes(color = rating), size = 1) +
  scale_color_brewer(type = 'seq', palette = "YlOrRd") +
  theme(panel.background = element_rect(fill = 'dark gray'))
```

These density plots show the trends of each feature by wine quality rating 
group. The most obvious trends being free SO2 proportion, volatile acidity, 
citric acid and alcohol.

```{r echo=FALSE}
ggplot(aes(x = fixed.acidity, y = pH, color = rating), data = wine) + 
  geom_point(size = 1, position = 'jitter') +
  scale_color_brewer(type = 'seq',
    guide = guide_legend(title = 'Quality', reverse = T,
    override.aes = list(alpha = 1, size = 2)))
```

fixed acidity, pH and quality do not appear to have any noticeable correlation. 
Premium, Good, and Poor wine groups seem to be similarly distrubted throughout 
the cluster of data ponits.

```{r echo=FALSE}
ggplot(aes(x = fixed.acidity, y = citric.acid, color = rating), data = wine) + 
  geom_point(size = 1, position = 'jitter') +
  scale_color_brewer(type = 'seq',
    guide = guide_legend(title = 'Quality', reverse = T,
    override.aes = list(alpha = 1, size = 2)))
```

There is a cluster of premium rated wines with more than 0.25 of citric acid and 
with more than 8 grams/cubic decimeter of fixed acidity.

```{r echo=FALSE}
ggplot(aes(x = fixed.acidity, y = log10(citric.acid + 1), color = rating), data = wine) + 
  geom_point(size = 1, position = 'jitter') +
  coord_cartesian(ylim = c(0, 0.25), xlim = c(5,12)) +
  scale_color_brewer(type = 'seq',
    guide = guide_legend(title = 'Quality', reverse = T,
    override.aes = list(alpha = 1, size = 2)))
```

Same graph as above with log of citric acid to lessen the effect of outliers.

```{r echo=FALSE}
ggplot(aes(x = fixed.acidity, y = density, color = rating), data = wine) + 
  geom_point(size = 1, position = 'jitter') +
  scale_color_brewer(type = 'seq',
    guide = guide_legend(title = 'Quality', reverse = T,
    override.aes = list(alpha = 1, size = 2)))
```

Higher quality wines tend to have lower density.

```{r echo=FALSE}
ggplot(aes(x = volatile.acidity, y = citric.acid, color = rating), data = wine) + 
  geom_point(size = 1, position = 'jitter') +
  scale_color_brewer(type = 'seq',
    guide = guide_legend(title = 'Quality', reverse = T,
    override.aes = list(alpha = 1, size = 2)))
```

There is a concentration of premium wines on the lower end of volatile acidity 
concentration and with citric acid between .25 and .75.

```{r echo=FALSE}
ggplot(aes(x = citric.acid, y = pH, color = rating), data = wine) + 
  geom_point(size = 1, position = 'jitter') +
  scale_color_brewer(type = 'seq',
    guide = guide_legend(title = 'Quality', reverse = T,
    override.aes = list(alpha = 1, size = 2)))
```

There is a cluster of premium wines with lower pH and higher citric acid 
concentration.

```{r echo=FALSE}
grid.arrange(
  
  ggplot(aes(x = alcohol, y = pH, color = rating), 
         data = wine) + 
  geom_point(size = 1, position = 'jitter') +
  scale_color_brewer(type = 'seq',
    guide = guide_legend(title = 'Quality', reverse = T,
    override.aes = list(alpha = 1, size = 2))),
  ggplot(aes(x = alcohol, y = density, color = rating), 
         data = wine) + 
  geom_point(size = 1, position = 'jitter') +
  scale_color_brewer(type = 'seq',
    guide = guide_legend(title = 'Quality', reverse = T,
    override.aes = list(alpha = 1, size = 2))))
```

These two scatter plots show what we know about higher rated wines having higher 
alcohol content. The trend does not seem in anyway influenced by the variables 
pH and density.

```{r echo=FALSE}
library(nnet)
library(stargazer)

# Checking the output (dependent variable)
table(wine$rating)
# Prepare data
mydata <- wine
mydata$freeSO2 <- (wine$free.sulfur.dioxide / wine$total.sulfur.dioxide)
mydata$out <- relevel(mydata$rating, ref = "Poor")
# Develop Multinomial Logistic regression Model
m1 <- multinom(out ~ volatile.acidity + alcohol + citric.acid + freeSO2, 
               data = mydata)
summary(m1)
```

Using this model based on the scatter plot, we can see that there is a decrease
in the log odds associated with the volatile acidity variable for both 
"Good" vs. "Poor" wine groups. 

An increase in log odds for the variable citric acid for every one unit 
increase.

The freeSO2 variable is the percentage of free sulfur dioxide compared with the
total amount of sulfur dioxide.

```{r}
# p-values
stargazer(m1, type = "text")
# Easier interpretation using relative risk ratios
m.rrr = exp(coef(m1))
m.rrr
```


Taking the exponential of the coefficients will allow for easier interpretation
of the logit coefficients.

For example keeping all other variables constant for every one unit increase of 
the variable "alcohol" it is 4.5 times more likely the wine will be in the
Premium group, where as for every one unit increase of "volatile.acidity" we see
the probability is less than 0 (0.003608785) times more likely for the wine to 
be in Premium group.

# Multivariate Analysis

### Talk about some of the relationships you observed in this part of the \
investigation. Were there features that strengthened each other in terms of \
looking at your feature(s) of interest?

Yes, Using scatter plots comparing features with quality rating as the color 
factor I was able to view clusters of wine groups. The best visualizations for
this where: Fixed acidity vs. citric acid, fixed acidity vs.density, 
volatile acidity vs. citric acid, and citric acid vs. pH. I knew from this that
I should try to use most of these features in any model.

### Were there any interesting or surprising interactions between features?
Yes I was surprised that citric acid and volatile acidity modeled wine quality
groups pretty well. Citric acid was more important than I thought it would be 
before starting the analysis.

### Did you create any models with your dataset? Discuss the \
strengths and limitations of your model.

I created a multinomial logit model. It is good for cases when the outcome 
variable is categorical with more than 2 levels.

The model was able to generate coeficients that supported my explorations of the
the data. Volatile acidity had a negative coefficient, and alcohol a positive 
one, for example. From this I was able to interpret how the model would use the
features in computing log odds for predictions.

The model however had a large Residual Deviance value of 2537.765, for a 
multinomial logit the deviance is preferred to be smaller. This could be that 
the features were not the best choice or there are still features that are not 
in the data that assist in creating a model.

------

# Final Plots and Summary

### Plot One

```{r echo=FALSE, Plot_One}
qplot(quality, data = wine, binwidth = 1) +
  scale_x_continuous(breaks = seq(3, 8, 1)) +
  ggtitle("Quality Distribution") + xlab("Quality") + ylab("Frequency")
```

### Description One

From this visualization it was observed that most of the wines were rated 
5 or 6. Very few other ratings were given relatively. Due to this the decision
was made to label wines with rating 3 to 5 as "Poor", 6 as "Good", and 7 to 8 
as "Premium" wines.

### Plot Two

```{r echo=FALSE, Plot_Two}
grid.arrange(
ggplot(aes(y = alcohol, x = rating, fill = rating), data = wine) +
  geom_boxplot() + coord_flip() + 
  labs(title = "Alcohol by Quality Rating") + 
  ylab(label = "Alcohol % by volume") + xlab("Quality Rating Group"),
ggplot(aes(y = volatile.acidity, x = rating, fill = rating), data = wine) +
  geom_boxplot() + coord_flip() + 
  labs(title = "Acetic acid by Quality Rating") +
  ylab(label = "Volatile Acidity (g/dm^3)") + xlab("Quality Rating Group")
)
```

### Description Two

Higher rated wines tend to have higher alcohol content and less acetic acid. I 
expected this trend because it was stated in data set text file that volatile
acidity can contribute to a strong vinegar taste in the wine. While higher 
alcohol seems to be an important characteristic to wines.

### Plot Three

```{r echo=FALSE, Plot_Three}
ggplot(aes(x = volatile.acidity, y = citric.acid, color = rating), data = wine) + 
  geom_point(size = 1, position = 'jitter') +
  scale_color_brewer(type = 'seq',
    guide = guide_legend(title = 'Quality', reverse = T,
    override.aes = list(alpha = 1, size = 2))) + 
  ggtitle("Citric Acid by Volatile Acidity and Quality Rating Grouping") +
  xlab("Volatile Acidity (Acetic Acid g / dm^3)") + ylab("Citric Acid") +
  coord_cartesian(ylim = c(0, 0.8), xlim = c(0.1,.8))
```

### Description Three

Premium wines cluster at higher citric acid content and lower acetic acid 
content.

------

# Reflection

The wine data set contains observations on nearly 1600 red wines, and initially
contained 13 features with no missing data. I explored the structure of the data
and identified the most important features and those that would not be important
to the analysis. I saw that the "Quality" were ratings made by wine tasting
experts. The minimum and maximum values were 3 and 8 respectively. I combined 
the the ratings into appropriate bins to make up for small amount of data for
certain ratings like, 3, 4 and 8. By combining the ratings into "Poor", "Good",
and "Premium" groups I was able to make data readings more legible.

The outcome variable Quality was most strongly correlated with Alcohol and 
Volatile Acidity features with pearson scores0.5 and -0.6 respectively. Of all
the pairs of correlations the strongest pearson scores were between 
Fixed Acidity vs. pH, Citric Acid, and Density. 

A limitation of this dataset is there are so few ratings except for 5 and 6, 
making me have to bin the data. I had to make the decision of which ratings 
would be poor and which would be best without the domain knowledge. I also think
this data could benefit from a machine learning algorithm and feature selection
to create a stronger model. Despite that with the multinomial logit model I 
identified features of significance (alcohol, volatile acidity, citric acid, 
and freeSo2).

